Computational methods in landscape ecology

### 2024-10-18 Sections/05-sampling.qmd
Further edits ✍️

Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
However, in practice, the appropriate scales may not be obvious or data at such scales may not be available.
Nevertheless, analyses at multiple spatial scales can help to identify the scales of interest.
For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape] or show the strongest relationships with response variables [@Huais2024MultilandrPackage].

@@

<!--
Software to perform multi-scale analysis include the R package *multilandr*, or the Python libraries *ZonalMetrics* [@Adamczyk2017ZonalMetricsPython], and *PyLandStats*.
-->

>>>>

Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
However, in practice, the appropriate scales may not be obvious or data at such scales may not be available and landscape-wide and local process interactions can occur across multiple scales [@miguet2016determines].
Here analyses at multiple spatial scales can help to identify the scales of interest. For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape] or show the strongest relationships with response variables [@Huais2024MultilandrPackage].

@@

### 2024-10-18 Sections/06-simulations.qmd
Further edits ✍️

Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
Thus, simulations can be seen as experimental systems that allow all imaginable manipulations which would be impossible in natural systems in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins]. 
The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g. drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
This is further facilitated by the increasing rate of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].

>>>>

Simulation models are a powerful tool to study complex adaptive socio-ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
Thus, simulations can be seen as experimental systems that allow all imaginable manipulations which would be impossible in natural systems in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling]. They can also incorporate a range of socio-economic, political and governance information to influence how agents (such as farmers) interact within a landscape [@murray2014combining].
In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins]. Digital twins are particular useful for real-time workflows that allow now- and forecasting of complex dynamics in a landscape.
The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g. drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
This is further facilitated by the increasing rate of novel sources of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].

Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins]. Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
